---
title: "Visual predictive checking with overlaid plots"
author: "Teemu Säilynoja"
date: "2023-01-19"
date-modified: last-modified
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    code-line-numbers: true
    default-image-extension: png
    fig-format: png
    embed-resources: true
knitr:
  opts_chunk:
    fig.path: ../../images/
    dev.args:
      bg: transparent
execute:
  cache: true
---

In this notebook we demonstrate the differences of quantile dot plots, histograms, and KDE density plots when overlaying predictive draws from the model on the observation.

```{r}
#| label: imports
#| message: false
library(ggplot2)
library(ggdist)
library(bayesplot)
library(cmdstanr)

source("code/R/helpers.R")
source("code/R/helpers_qdotplot.R")
source("code/R/helpers_kde.R")
source("code/R/helpers_histogram.R")

theme_set(ppc_paper_theme(28) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.line.y = element_blank()
  ))

bayesplot_theme_set(ppc_paper_theme(28) +
  theme(
    axis.title = element_blank(),
    legend.position = "none",
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.line.y = element_blank()
  ))

color_scheme_set(unname(paper_colors[1:6]))
set.seed(9865875)
```

We start by creating a normally distributed sample.
```{r}
#| label: make_data
n <- 1e3
nq <- 100
bw_dots <- sqrt(1 / nq)
y <- rnorm(n, sd = .4)
```

We fit the following Stan model to the sample and extract predictive draws.

```{r}
#| label: show_stan_model
#| class-output: stan

model <- cmdstan_model(stan_file = "code/stan/02_05_normal.stan")

model$print()
```
```{r}
#| label: fit_stan_model

fit <- model$sample(data = list(N = n, Y = y), refresh = 0, thin = 50)
```

```{r}
#| label: bin_yrep
yrep <- fit$draws(variables = "yrep", format = "matrix")
## For histogram binning:
yrep_h <- yrep
yrep_h[yrep_h < min(y)] <- min(y)
yrep_h[yrep_h > max(y)] <- max(y)
```


## PPC with quantile dot plots

We plot the observation as quantile dot plot with 100 quantiles. Then, we overlay just the top dot of each stack to visualize the variation in the quantile dot plots of the predictive draws. As the dots from the same draw are not linked in the visualization, we can't assess the variation in the overall shape between draws. 
```{r}
#| label: 02_05_qdot_overlay
qdots <- ggplot(
  bin_dots(
    x = quantile(y, stats::ppoints(nq, a = 1 / 2), type = 5),
    y = 0,
    binwidth = bw_dots
)) +
  ggforce::geom_circle(
    aes(x0 = x, y0 = y, r = .5 * bw_dots),
    alpha = .2,
    fill = paper_colors["light_highlight"],
    data = apply(
      yrep[1:50,],
      1,
      \(row) bin_dots(
        x = quantile(row, stats::ppoints(nq, a = 1 / 2), type = 5),
        y = 0,
        binwidth = bw_dots
      ) |>
        dplyr::group_by(bin) |>
        dplyr::summarise(x = min(x), y = max(y)) |>
        dplyr::select(x, y, bin)
    ) |> dplyr::bind_rows(),
    linewidth = 0,
    n = 10
  ) +
    ggforce::geom_circle(
    aes(x0 = x, y0 = y, r = 0.5 * bw_dots),
    color = "black",
    linewidth = .5
  ) +
  coord_equal()

qdots
```

## PPC with histogram

We compute the bin width with the observation and use this and the bin breaks also for visualizing the posterior predictive draws.
```{r}
#| label: binwidth_computation
bw_hist <- 2 * IQR(y) / length(y)^(1 / 3)
```

Overlaying the predictive mean and credible interval of each bin only allows us to compare the predictions to the observation per individual bin. Like with quantile dot plots, we can't see the variation in hte overall shape of the distribution between the predictive draws.
```{r}
#| label: 02_05_hist_overlay

p_hist <- ggplot(data.frame(y = y)) +
  geom_histogram(
    aes(
      x = y,
      y = after_stat(density)
    ),
    fill = "white", #paper_colors["light_highlight"],
    color = paper_colors["dark_highlight"],#paper_colors["mid"],
    linewidth = .6,
    binwidth = bw_hist,
    center = mean(range(y))
  )

p_hist <- p_hist +
  geom_pointinterval(
    data = yrep_h |> apply(1, \(row) hist(
      row,
      breaks = c(layer_data(p_hist)$xmin[1], layer_data(p_hist)$xmax),
      plot = F
    )[c("mids", "density")]) |> dplyr::bind_rows() |> dplyr::group_by(mids) |> dplyr::summarise(
      y = mean(density),
      ymin = quantile(density, .05),
      ymax = quantile(density, .95)
    ),
    aes(
      x = mids,
      y = y,
      ymin = ymin,
      ymax = ymax
    ),
    color = paper_colors["mid"],#paper_colors["dark_highlight"],
    interval_color = paper_colors["mid"], #paper_colors["dark"],
    linewidth = 5.0,
    size = 5.0
  ) +
  coord_equal(
    xlim = layer_scales(qdots)$x$range$range,
    ylim = layer_scales(qdots)$y$range$range
  )

p_hist
```

## PPC with overlaid KDE density plots

`bayesplot` already has overlaid KDE densities implemented. This is a common visual PPC plot and shows an overview of the observation and draws from the predictive distribution.

```{r}
#| label: 02_05_kde_overlay

ppc_dens_overlay(y, yrep, alpha = 1) +
  coord_equal(
    xlim = layer_scales(qdots)$x$range$range,
    ylim = layer_scales(qdots)$y$range$range
  )
```



